---
title: "Density estimation, clustering, PCA and
principal curves"
author: "Gonzalo Solera, Zofia Tarant"
date: "11 12 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r data reading}
data <- read.table('zip.train')
# get only zeros
data <- data[which(data[1]==0), -1]
```
```{r plot function, echo=FALSE}
plot.zip <- function(x){
  x<-as.numeric(x)
  x.mat <- matrix(x,16,16)
  image(1:16,1:16,x.mat[,16:1], col=gray(seq(1,0,l=12)))
}
```

## Hierarchical clustering

Euclidean distance:
```{r Euclidean distance}
d.Eucl <- dist(data)
summary(d.Eucl)
```

Clustering:
```{r hierarchical clustering}
k <- 4
hcl.Eucl <- hclust(d.Eucl, method="ward.D")
plot(hcl.Eucl)
hcl.Eucl.k <- cutree(hcl.Eucl,k)
table(hcl.Eucl.k)

clusters <- hcl.Eucl.k

# Plot the average digit of each cluster
op<-par(mfrow=c(2,2))
for(i in 1:k) {
  mean.digit <- colMeans(data[which(clusters == i), ])
  plot.zip(mean.digit)
  title(paste("cluster", i, sep=" "))
}

```
## Principal components

```{r PCs}
principal_components_obj <- princomp(data)
summary(principal_components_obj)
```
```{r scatterplot of PC}
principal_components <- principal_components_obj$scores[, 1:2]
plot(principal_components, pch=19, cex=.5, col=clusters)
```

## Density estimation

4. For each one of the k clusters obtained above, do the following tasks:
• Consider the bivariate data set of the scores in PC1 and PC2 of the points in this cluster.
• Estimate non-parametrically the joint density of (PC1,PC2), conditional to this cluster. Use the default
bandwith values.
• Represent the estimated bivariate density using the level curve that covers the 75% of the points in this
cluster.
A unique plot should be done, at which the k densities are represented simultaneously.
5. Over the previous plot, represent the principal curve obtained using the package princurve.

```{r density estimation of PC}
library('ks')
# clustered.PC$scores[,1:2]

kde_1 = ks::kde(x = principal_components[which(clusters==1), 1:2])
kde_2 = ks::kde(x = principal_components[which(clusters==2), 1:2])
kde_3 = ks::kde(x = principal_components[which(clusters==3), 1:2])
kde_4 = ks::kde(x = principal_components[which(clusters==4), 1:2])

plot(kde_1, cont=75, col=1, xlim=c(-10, 10), ylim=c(-10, 10))
plot(kde_2, cont=75, col=2, add=TRUE)
plot(kde_3, cont=75, col=3, add=TRUE)
plot(kde_4, cont=75, col=4, add=TRUE)


library(princurve)
principal_curve <- principal_curve(as.matrix(data))
principal_curve_projection <- predict(principal_components_obj, principal_curve$s)

lines(principal_curve_projection[principal_curve$ord, 1:2], col=4, lwd=2)
```

6. For each one of the k clusters obtained above, do the following tasks:
• Consider the univariate data set of the lambda scores over the principal curve of the points in this
cluster.
• Estimate non-parametrically the density function of lambda, conditional to this cluster. Use the default
bandwith value.
• Plot the estimated density function.
A unique plot should be done, at which the k densities are represented simultaneously.

```{r density estimation of principal curve lambda}
kde_lambda_1 = ks::kde(x = principal_curve$lambda[which(clusters==1)])
kde_lambda_2 = ks::kde(x = principal_curve$lambda[which(clusters==2)])
kde_lambda_3 = ks::kde(x = principal_curve$lambda[which(clusters==3)])
kde_lambda_4 = ks::kde(x = principal_curve$lambda[which(clusters==4)])

plot(kde_lambda_1, col=1, xlim=c(-5, 40), ylim=c(0, 0.2), xlab='lambda')
plot(kde_lambda_2, col=2, add=TRUE)
plot(kde_lambda_3, col=3, add=TRUE)
plot(kde_lambda_4, col=4, add=TRUE)
```
